Adding Sensitivity to 21cm Inteferometric Probes of Reionization by Optimizing Choice of Baselines

### 2016-11-16 Rotation paper/draft_working.tex
update

\altaffiltext{1}{Astronomy Dept., U. California, Berkeley, CA}
\altaffiltext{2}{Radio Astronomy Lab., U. California, Berkeley, CA}
%\altaffiltext{3}{Berkeley Center for Cosmological Physics, Berkeley, CA}
%\altaffiltext{3}{Dept. of Physics and Astronomy, U. Pennsylvania, Philadelphia, PA}
%\altaffiltext{8}{School of Earth and Space Exploration, Arizona State U., Tempe, AZ}

\begin{abstract}
We introduce an accurate method to extract additional sensitivity from data of redundant radio arrays. 
Using the 64 element version of Precision Array for Probing the Epoch of Reionization (PAPER-64), we illustrate 
how our method finds optimal baselines to cross correlate, including baselines that are slightly different in length and orientation. 
We predict the sensitivity improvements of including each baseline pairs. The extremely natural and simple implementation of this method
would increase the sensitivity of PAPER-64 by up to [~50] percent. 
\end{abstract}

@@

One of the main challenges of interferometric observations of cosmic
21cm line is the foreground contamination. Both sources within our
galaxy, and to a lesser distant sources outside of our galaxy emit
radio contaminations (via for example synchrotron processes) up to five orders of magnitude stronger than the
reionization signal. There are two common methods to deal with the
foreground contamination. In the first, sometimes called the foreground
removal technique, individual sources are identified and removed in the image domain.
The other technique, commonly referred to as 'foreground avoidance',
makes use of the fact that most common foreground contaminations have a smooth
spectrum, and thus is constrained to a Fourier domain ``foreground
wedge'' \cite{wedge1,wedge2}. Thus contamination can be avoided
by restricting our observation to ``outside the wedge'', where the data is currently noise dominated. The main
challenge of using the avoidance technique is thus the sensitivity, or
the signal to noise ratio. This is the motivation for the design of
the maximum redundancy arrays such as PAPER and HERA. While traditional
imaging arrays focus on Fourier mode uv coverage, the redundant arrays
have many tightly packed antennas at equal spacing. Since baselines
of the same length and orientation measure the same Fourier modes
on the sky, the maximum redundancy arrays is able to increase the
signal to noise ratio by averaging over measurements of the same baseline.
\cite{Ali2015} provides the newest upper limit to the power spectrum
measurements with the PAPER-64 data. From some models we roughly expect
that the sensitivity required for detection is a factor of 10 away.
This paper explores a new technique to add sensitivity to the analysis
of redundant arrays, by making use of the earth's rotation. This technique
can be applied to redundant arrays such as PAPER and HERA to increase
the sensitivity by up to a factor of [1.5].

@@

\subsection{Rough Idea}
\begin{figure}[H]
\includegraphics[width=\linewidth]{64layout}

\caption{The PAPER-64 layout. Each blue dot corresponds to the location of
an antenna. Top panel shows the antenna positions drawn to scale,
bottom panel show the antenna labels and distances.\label{fig:The-PAPER-64}
The numbering of the antennas the bottom panel are original labels
during instrument assembly and does not bear significant
meaning. Baselines corresponding to the same separations are called
equivalent. In the bottom panel, the two baselines labeled with the
red arrow are equivalent, with separation denoted by sep1,0, for the
antennas are separated by 1 unit east and 0 unit north. Similarly,
the baselines labeled with the yellow and green arrows are examples
of sep1,1 and sep-1,1, respectively. Note sep1,0 and sep-1,0 for example
are the same baselines and should not be counted twice. }
\end{figure}

We shall use the 64-antenna PAPER array to demonstrate. 
The PAPER-64 array is located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern with antenna labels are show
in Fig. \ref{fig:The-PAPER-64}. We see the antenna spacing in North-South
directions are comparatively close (4m), so that baselines such as
0\_44 and 0\_7 are very close to equivalent. In the bottom panel,
the two baselines labeled with the red arrow are equivalent, with
separation denoted by sep1,0, for the antennas are separated by 1
unit east and 0 unit north. Similarly, the baselines labeled with
the yellow and green arrows are examples of sep1,1 and sep-1,1, respectively.
Note sep1,0 and sep-1,0 for example are the same baselines and should
not be counted twice. Turns out antennas in purely north-south baselines
are close enough to induce cross talks, and hence are not suitable
for use. The original PAPER-64 analysis thus used three classes of baselines,
sep1,0, sep1,1 and sep-1,1 \cite{Ali2015}. There each of these classes
of baselines are cross multiplied to itself. We shall see that in addition these
baseline classes can be cross multiplied with a time offset.

Given a point source on the sky, each baseline maps the
source to a point in uv plane. As the earth rotates with respect to
the source, the point traces out tracks in the uv-plane. 
The uv tracks of PAPER-64 over 7.2 hours, at 0.15GHz, for a source that passes through zenith, are shown in Fig. \ref{fig:Tracks-of-PAPER-64}.

\begin{figure}[H]
\includegraphics[width=\linewidth]{TRACKSpt3days}
\caption{Tracks of PAPER-64 for a hypothetical source that passes through zenith.
These tracks are traced out over 0.3 sidereal days, or roughly 7.2
hours. Color represents different baselines. Frequency is $\nu=0.15\text{GHz}$.
As the earth rotates, tracks are traced out counterclockwise. \label{fig:Tracks-of-PAPER-64}}
\end{figure}

Roughly speaking, we can identify
redundancy of near-equivalent baselines as crossings
of the uv tracks. Although this is a valid quick method to identify some baselines to cross-multiply, we find
that it is not accurate enough for time offset determination. The reason is, for wide beam antennas such as PAPER-64, the sources are not point-like. 
In other words, when tracks of
one point of the sky are crossing in uv-plane, those corresponding to another is not. To accurately
constrain redundancy we must take into account of finite beam width. A more rigorous treatment is thus required.  
We show the beam of PAPER-64 in Fig.\ref{fig:Beam-of-PAPER-64.} for reference:

@@

\caption{Beam of PAPER-64. Shown here is the values of $A$ (Eq. \eqref{eq:Vis1})
for frequency of $\nu=150\text{MHz}$. Notice $A$ is a ``baseline's
beam''. Since all antennas in PAPER-64 have the same beam, $A$ here
is just the square of an antenna beam, and is normalized to 1 at the
center. The circles centered around zenith (center of beam) here are
spaced 10 degrees apart. Such a wide beam limits the accuracy of the uv track-crossing as a method of redundancy search. \label{fig:Beam-of-PAPER-64.}}
\end{figure}

@@

>>>>

\altaffiltext{1}{Astronomy Dept., U. California, Berkeley, CA}
\altaffiltext{2}{Radio Astronomy Lab., U. California, Berkeley, CA}

\begin{abstract}
We introduce an accurate method to extract additional sensitivity from data of redundant radio arrays. 
Using the latest and final 128-element version of Precision Array for Probing the Epoch of Reionization (PAPER-64), we illustrate 
how our method finds optimal baselines to cross correlate, including baselines that are slightly different in length and orientation. 
We predict the sensitivity improvements of including each baseline pairs. The extremely natural and simple implementation of this method
would increase the sensitivity of PAPER-128 by up to [~50] percent. 
\end{abstract}

@@

One of the main challenges of interferometric observations of cosmic
21cm line is the foreground contamination. Both sources within our
galaxy, and to a lesser distant sources outside of our galaxy emit
radio contaminations (via for example synchrotron processes) up to five orders of magnitude stronger than the
reionization signal. There are two common methods to deal with the
foreground contamination. In the first, sometimes called the foreground
removal technique, individual sources are identified and removed in the image domain.
The other technique, commonly referred to as 'foreground avoidance',
makes use of the fact that most common foreground contaminations have a smooth
spectrum, and thus is constrained to a Fourier domain ``foreground
wedge'' \cite{wedge1,wedge2}. Thus contamination can be avoided
by restricting our observation to ``outside the wedge'', where the data is currently noise dominated. The main
challenge of using the avoidance technique is thus the sensitivity, or
the signal to noise ratio. This is the motivation for the design of
the maximum redundancy arrays such as PAPER and HERA. While traditional
imaging arrays focus on Fourier mode uv coverage, the redundant arrays
focus on redundant measurements of the same Fourier mode
with many tightly packed antennas at equal spacing. Since baselines
of the same length and orientation measure the same Fourier modes
on the sky, the maximum redundancy arrays is able to increase the
signal to noise ratio by averaging over measurements of the same baseline.
\cite{Ali2015} provides the newest upper limit to the power spectrum
measurements with the PAPER-64 data. From some models we roughly expect
that the sensitivity required for detection is a factor of 10 away.
This paper explores a new technique to add sensitivity to the analysis
of redundant arrays, by making use of the earth's rotation. This technique
can be applied to redundant arrays such as PAPER and HERA to increase
the sensitivity by up to a factor of [1.5].

@@

\subsection{Rough Idea}
\begin{figure}[H]
\includegraphics[width=\linewidth]{antpos128}

\caption{The PAPER-128 layout. Each blue dot corresponds to the location of
an antenna. Top panel shows the antenna positions drawn to scale,
bottom panel show the antenna labels and distances, excluding the outlier
antennas.\label{fig:AntPos}
The numbering of the antennas the bottom panel are original labels
during instrument assembly and does not bear significant
meaning. Baselines corresponding to the same separations are called
equivalent. In the bottom panel, the two baselines labeled with the
red arrow are equivalent, with separation denoted by sep1,0, for the
antennas are separated by 1 unit east and 0 unit north. Similarly,
the baselines labeled with the yellow and green arrows are examples
of sep1,1 and sep-1,1, respectively. Note sep1,0 and sep-1,0 for example
are the same baselines and should not be counted twice.}
\end{figure}

We shall use the 128-element PAPER array to demonstrate. 
The PAPER array is located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern with antenna labels are show
in Fig. \ref{fig:AntPos}. We see the antenna spacing in North-South
directions are comparatively close (4m), so that baselines such as
0\_44 and 0\_7 are very close to equivalent. In the bottom panel,
the two baselines labeled with the red arrow are equivalent, with
separation denoted by sep1,0, for the antennas are separated by 1
unit east and 0 unit north. Similarly, the baselines labeled with
the yellow and green arrows are examples of sep1,1 and sep-1,1, respectively.
Note sep1,0 and sep-1,0 for example are the same baselines and should
not be counted twice. Turns out antennas in purely north-south baselines
are close enough to induce cross talks, and hence are not suitable
for use. The original PAPER-64 analysis used three classes of baselines, here
equivalent to 
sep2,0, sep2,1 and sep-2,1 \cite{Ali2015}. There each of these classes
of baselines are cross multiplied to itself. We shall see that in addition these
baseline classes can be cross multiplied with a time offset.

Given a point source on the sky, each baseline maps the
source to a point in uv plane. As the earth rotates with respect to
the source, the point traces out tracks in the uv-plane. 
The uv tracks of PAPER-128 over 7.2 hours, at 0.15GHz, for a source that passes through zenith, are shown in Fig. \ref{fig:Tracks}.

\begin{figure}[H]
\includegraphics[width=\linewidth]{tracks128}
\caption{Tracks of PAPER-128 (grid only, excluding outliers) for a hypothetical source that passes through zenith.
These tracks are traced out over 0.2 sidereal days, or roughly 4.8
hours. Color represents different baselines. Frequency is $\nu=0.15\text{GHz}$.
As the earth rotates, tracks are traced out counterclockwise. \label{fig:Tracks}}
\end{figure}

Roughly speaking, we can identify
redundancy of near-equivalent baselines as crossings
of the uv tracks. Although this is a valid quick method to identify some baselines to cross-multiply, we find
that it is not accurate enough for time offset determination. The reason is, for wide beam antennas such as PAPER, the sources are not point-like. 
In other words, when tracks of
one point of the sky are crossing in uv-plane, those corresponding to another is not. To accurately
constrain redundancy we must take into account of finite beam width. A more rigorous treatment is thus required.  
We show the beam of PAPER antennas in Fig.\ref{fig:Beam} for reference:

@@

\caption{Beam of PAPER antennas. Shown here is the values of $A$ (Eq. \eqref{eq:Vis1})
for frequency of $\nu=150\text{MHz}$. Notice $A$ is a ``baseline's
beam''. Since all antennas in PAPER have the same beam, $A$ here
is just the square of an antenna beam, and is normalized to 1 at the
center. The circles centered around zenith (center of beam) here are
spaced 10 degrees apart. Such a wide beam limits the accuracy of the uv track-crossing as a method of redundancy search. \label{fig:Beam}}
\end{figure}

@@

\subsection{Validation with PAPER-128 Data}
